---
title: "NCRI Octocoral Host Phylogenies"
author: "Matias Gomez"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_depth: 4
    toc_float: true
---

# 1. Overview

This document performs mitochondrial mtMutS phylogenetic inference using IQ TREE and visualization using ggtree for:

Briareum asbestinum

Eunicea caribaeorum

Eunicea flexuosa

Muricea muricata

# 2. Load R Packages

```{r}
if (!requireNamespace("BiocManager", quietly = TRUE))
install.packages("BiocManager")

required_pkgs <- c("ggtree", "ape", "ggplot2", "patchwork")

for (pkg in required_pkgs) {
if (!require(pkg, character.only = TRUE)) {
BiocManager::install(pkg, ask = FALSE)
library(pkg, character.only = TRUE)
}
}
```

# 3. Define Base Directory (EDIT THIS ON YOUR MACHINE)

This single line controls all file paths:

```{r}
base_dir <- "~/Desktop/NCRI_Octocoral_ID_2025/Octocoral_Host_ID_Clean_Sequences/mtMuts"
```


# 4. Function for Loading, Rooting, Renaming, and Plotting Trees

```{r}
process_tree <- function(tree_file, annotation_file, outgroup, title_expr, xlim_val) {

tree <- read.tree(tree_file)
tree_rooted <- root(tree, outgroup = outgroup, resolve.root = TRUE)

annotation <- read.delim(annotation_file, header = TRUE, stringsAsFactors = FALSE)

tree_rooted$tip.label <- annotation$new_label[
match(tree_rooted$tip.label, annotation$old_label)
]

tree_rooted$node.label[1] <- NA

p <- ggtree(tree_rooted, size = 1.2) +
geom_tippoint(shape = 21, size = 3, fill = "black") +
geom_tiplab(size = 7, hjust = -0.1, fontface = "italic") +
geom_nodelab(aes(label = label), size = 8, hjust = 2, vjust = 0) +
theme_tree2() +
ggtitle(title_expr) +
xlab("Number of nucleotide substitutions") +
xlim(0, xlim_val) +
theme(
axis.text.x = element_text(size = 18),
axis.title.x = element_text(size = 18, face = "bold"),
plot.title = element_text(size = 20, face = "bold", hjust = 0.5)
)

return(p)
}

```


# 5. Generate All Four Trees

```{r, eval=TRUE}
BAST_tree <- process_tree(
tree_file = file.path(base_dir, "BAST/BAST_mtMuts_IQTREE_Output/BAST_mtMuts_IQtree.contree"),
annotation_file = file.path(base_dir, "BAST/BAST_mtMuts_Alignments/BAST_mtMuts_Annotation_file.txt"),
outgroup = "NC_062000.1",
title_expr = expression(bolditalic("B. asbestinum") ~ bold(" mtMutS (801 bp) IQ TREE tree")),
xlim_val = 0.005
)

BAST_tree

ECAR_tree <- process_tree(
tree_file = file.path(base_dir, "ECAR/ECAR_mtMuts_IQTREE_Output/ECAR_mtMuts_IQtree.contree"),
annotation_file = file.path(base_dir, "ECAR/ECAR_mtMuts_Alignments/ECAR_mtMuts_Annotation_file.txt"),
outgroup = "NC_073493.1",
title_expr = expression(bolditalic("E. caribaeorum") ~ bold(" mtMutS (744 bp) IQ TREE tree")),
xlim_val = 0.16
)

ECAR_tree

EFLE_tree <- process_tree(
tree_file = file.path(base_dir, "EFLE/EFLE_mtMuts_IQTREE_Output/EFLE_mtMuts_IQtree.contree"),
annotation_file = file.path(base_dir, "EFLE/EFLE_mtMuts_Alignments/EFLE_mtMuts_Annotation_file.txt"),
outgroup = "GQ342524.1",
title_expr = expression(bolditalic("E. flexuosa") ~ bold(" mtMutS (734 bp) IQ TREE tree")),
xlim_val = 0.01
)

EFLE_tree

MMUR_tree <- process_tree(
tree_file = file.path(base_dir, "MMUR/MMUR_mtMuts_IQTREE_Output/MMUR_mtMuts_IQtree.contree"),
annotation_file = file.path(base_dir, "MMUR/MMUR_mtMuts_Alignments/MMUR_mtMuts_Annotation_file.txt"),
outgroup = "OZ247669.1_17312-18118",
title_expr = expression(bolditalic("M. muricata") ~ bold(" mtMutS (766 bp) IQ TREE tree")),
xlim_val = 0.05
)

MMUR_tree

```

# 6. Merge into 2 × 2 Panel

```{r}
mtMuts_panel <- (BAST_tree + ECAR_tree) /
(EFLE_tree + MMUR_tree) +
plot_annotation(
tag_levels = "a",
tag_suffix = ")",
theme = theme(
plot.tag = element_text(size = 40, face = "bold"),
plot.margin = margin(20, 20, 20, 20)
)
) & theme(
plot.tag.position = c(0, 1),
plot.margin = margin(20, 20, 20, 20)
)

mtMuts_panel

```

# 7. Save Final Figure

```{r, eval=FALSE}
ggsave(
"mtMuts_4panel_final.pdf",
mtMuts_panel,
width = 90,
height = 60,
units = "cm",
dpi = 600,
limitsize = FALSE,
bg = "white"
)

```

